---
title: "Exploratory_analysis"
author: "Biswajit Chowdhury"
date: "28/07/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- Here, I used the Heart disease data set from UCI machine learning repository. The dataset contains 14 variables and 303 observations.

# Load library
```{r}
library(tidyverse)

```

# Reading data file

```{r}
heart_data<-read.csv("heart.csv")

# see the data set
head(heart_data) # first 6 rows of the data set
tail(heart_data) # last 6 rows of the data set

dim(heart_data) # no. of rows and columns

colnames(heart_data) # variable names

# change the column names 
names(heart_data) <- c("age", "sex", "chest_pain", "resting_bp","cholestrol",
                       "fasting_sugar", "resting_ECG", "max_heart_rate", 
                       "exercise_agina", "oldpeak", "slope", "number_major_vessels",
                       "thal", "disease_status")


```


# preprocessing data for the exploratory analysis

```{r}
# Variable sex is coded 0 and 1. We want to include 0=F and 1=M

heart_data$sex <- factor(heart_data$sex,
      levels = c(0,1),
      labels = c("F", "M"))

# Change variable chest_pain into four categories
heart_data$chest_pain <- factor(heart_data$chest_pain,
      levels = c(1,2,3,0),
      labels = c("Typical_angina", "Atypical_angina", "Nonanginal_pain", "Assymptomatic"))

# Change variable fasting_sugar to high and low

heart_data$fasting_sugar <- factor(heart_data$fasting_sugar,
      levels = c(0,1),
      labels = c("Low", "High"))

# Change resting_ECG into normal, mild and severe

heart_data$resting_ECG <- factor(heart_data$resting_ECG,
      levels = c(0,1,2),
      labels = c("Normal", "Mild", "Severe"))

# Change excercise_angina into "yes" and "no"

heart_data$exercise_agina <- factor(heart_data$exercise_agina,
      levels = c(0,1),
      labels = c("No", "Yes"))

# Similary we also change the labels for number of major vessels and thal


heart_data$number_major_vessels <- factor(heart_data$number_major_vessels,
      levels = c(0,1,2,3,4),
      labels = c("None", "One", "Two", "Three", "Four"))

heart_data$thal <- factor(heart_data$thal,
      levels = c(1, 2,3),
      labels = c("Normal", "Fixed_defect", "Reversible_defect"))

# Finally change the angiographic disease status to normal and disease

heart_data$disease_status <- factor(heart_data$disease_status,
      levels = c(0,1),
      labels = c("Normal", "Disease"))

write.csv(heart_data, "UCI_heart_data.csv")
head(heart_data,3)
```

# Explore the quantitative variables

# Histogram and density plot for the quantitative variables 
```{r}
par(mfrow=c(3,2))
hist(heart_data$age,
     col="white",
     border="black",
     prob = TRUE,
     xlab = "age",
     main = "distribution of age")
  lines(density(heart_data$age),
      lwd = 2,
      col = "red")

hist(log(heart_data$age),
     col="white",
     border="black",
     prob = TRUE,
     xlab = "age",
     main = "distribution of log(age)")
  lines(density(log(heart_data$age)),
      lwd = 2,
      col = "red")
  
hist(heart_data$resting_bp,
     col="white",
     border="black",
     prob = TRUE,
     xlab = "resting_bp (mmHg)",
     main = "distribution of resting_bp")
  lines(density(heart_data$resting_bp),
      lwd = 2,
      col = "red")

hist(log(heart_data$resting_bp),
     col="white",
     border="black",
     prob = TRUE,
     xlab = "resting_bp (mmHg)",
     main = "distribution of log(resting_bp)")
  lines(density(log(heart_data$resting_bp)),
      lwd = 2,
      col = "red")
  
hist(heart_data$cholestrol,
     col="white",
     border="black",
     prob = TRUE,
     xlab = "cholestrol levels (mg/dl)",
     main = "distribution of cholestrol levels")
  lines(density(heart_data$cholestrol),
      lwd = 2,
      col = "red")
 
 hist(log(heart_data$cholestrol),
     col="white",
     border="black",
     prob = TRUE,
     xlab = "cholestrol levels (mg/dl)",
     main = "distribution of log(cholestrol levels)")
  lines(density(log(heart_data$cholestrol)),
      lwd = 2,
      col = "red")

  hist(heart_data$max_heart_rate,
     col="white",
     border="black",
     prob = TRUE,
     xlab = "max_heart_rate ",
     main = "distribution of max_heart_rate")
  lines(density(heart_data$max_heart_rate),
      lwd = 2,
      col = "red") 
  
  hist(log(heart_data$max_heart_rate),
     col="white",
     border="black",
     prob = TRUE,
     xlab = "max_heart_rate ",
     main = "distribution of max_heart_rate")
  lines(density(log(heart_data$max_heart_rate)),
      lwd = 2,
      col = "red") 
```


# Shapiro-Wilk test for checking the normal distribution
- null hypothesis: the data are normally distributed
- alternative hypothesis: the data are not normally distributed


```{r}
# normality test for cholestrol
shapiro.test(heart_data$cholestrol)

# normality test for heart ratel
shapiro.test(heart_data$max_heart_rate)
```
 - from the above p-value, we can say the data are not normally distributed.
 
## Visual inspection of the data normality using Q-Q plots (quantile-quantile plots). Q-Q plot draws the correlation between a given sample and the normal distribution.

```{r}
library(ggpubr)
ggqqplot(heart_data$cholestrol, ylab="Cholestrol")

ggqqplot(heart_data$max_heart_rate, ylab="Maximum heart rate")
```

## Pearson correlation test

- Correlation between cholestrol and heart_rate

```{r}
# we can use three different methods: pearson, kendal, spearman
correlation<-cor.test(heart_data$cholestrol, heart_data$max_heart_rate, method="pearson")
correlation
```

p-value is greater than the significance level alpha =0.05. Therefore, we can conclude that above two variables are not significantly correlated with a correlation coefficiengt of -0.009 and the p-value of 0.8632.
```{r}
library(ggpubr)

# Presence of angiographic disease status among males and females due to age  

ggboxplot(heart_data, x="disease_status", y="age",
          xlab="angiographic disease status", ylab= "Age (years)",
          title= "disease_status with age and sex",
          color = "sex", # color by gender
          add = c("jitter", "mean_sd")) # show the distribution of the observed values


# Presence of angiographic disease status among males and females due to resting blood pressue  

ggboxplot(heart_data, x = "disease_status", y = "resting_bp", 
          xlab = "Presence of angiographic disease", ylab = "Resting blood pressue (mm Hg)", 
          title = "Resting blood pressue and angiographic disease status",
          color = "sex", add = c("jitter", "mean_sd"))
```


```{r}
#par(mfrow=c(1,2))
# Changes of chlestreol level between angiographic disease status and normal people  

ggboxplot(heart_data, x = "disease_status", y = "cholestrol", 
          xlab = "Presence of angiographic disease", ylab = "Cholestoral level (mg/dl)",
          title = "Relationship between angiographic disease status and serum cholestoral",
          color = "sex", add = c("jitter", "mean_sd")) 

# Chnages of heart rate between normal and angiographic disease 

ggboxplot(heart_data, x = "disease_status", y = "max_heart_rate", 
          xlab = "Presence of angiographic disease", ylab = "Heart rate", 
          title = "Relationship between angiographic disease status and heart rate",
          color = "sex", add = c("jitter", "mean_sd")) 
```




# Explore the categorical variables

- Calculate the frequency of categorical variables

```{r}
# Counts for the sex categories

table(heart_data$sex)

# cross classification for sex with angiographic disease status
table(heart_data$disease_status, heart_data$sex)

```

- Assess subjects by disease type, gender, fasting sugar and chest pain

```{r}
# Multidimensional tables based on three or more categorical variables

table1 <- table(heart_data$disease_status, heart_data$sex, 
                 heart_data$fasting_sugar)

ftable(table1) # print the results with more attractively
  
table2 <- table(heart_data$disease_status, heart_data$sex, 
                 heart_data$chest_pain)

ftable(table2)
```

# Visualization of the categorical variables

```{r}
# Here we assess the number of subjects diagnosed with chest pain
ggplot(heart_data, aes(x = chest_pain, fill=chest_pain)) +
        geom_bar() +  xlab('Level of chest pain') + ylab("Count") 

```

- To easy to digest of the above image, we can make it in desending order. For this we will make a functions that sort the variables crosponding their total counts

```{r}
# re-order levels
reorder_size <- function(x) {
        factor(x, levels = names(sort(table(x), decreasing = TRUE)))
}

ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar() +
  xlab('Level of chest pain') + ylab("Count")
```

# Changes of chest pain with sex and disease status

```{r}
ggplot(heart_data, aes(x = reorder_size(chest_pain), fill=chest_pain)) +
        geom_bar() +
  xlab('Level of chest pain') +
  facet_grid(disease_status~ sex)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Number of major vessels with gender and heart disease

```{r}
ggplot(heart_data, aes(x = reorder_size(number_major_vessels), fill=number_major_vessels)) +
        geom_bar() +
  xlab('Number of major vessels') +
  facet_grid(disease_status~ sex) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Proportion/Percentage

```{r}
# percentages of gender categories
table1<- table(heart_data$sex)
prop.table(table1)

# percentage of cross classication counts for gender by disease types

table2<- table(heart_data$disease_status, heart_data$sex)
prop.table(table2)

round(prop.table(table2), 3)*100
```

# co-relation plot between quantitative variables

```{r}
library(corrplot)

corrplot(cor(heart_data[, c(1,4,5,8,10)]), type="upper")
```

